JD notes on effects

### 2022-06-15 manuscript/body.tex
Autocommit

@@

\item \textbf{Anchor:} The value chosen for the focal input variable when estimating effect confidence intervals. The anchor choice does not affect the main prediction estimates. Typically chosen as the center point of the predictor variables corresponding to the focal input variable but other sensible values can be chosen too. If we are trying to focus on effects, the anchor from which we calculate CIs becomes important. This is not true for the more conventional prediction plots: we can choose any anchor, e.g., zero-anchored, but model center seems like a natural, stable choice, i.e., center-anchored. In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal input variable.

@@

To get an intuition of how conditioning on the mean values of non-focal predictors work, suppose we are interested in predictions at the levels of a particular predictor, i.e., focal, $x_f$, from the set of predictors. As a starting point, assume that the model has no interaction terms. The idea is to choose a reference point for the values of non-focal predictors. For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors. We refer to this as centered or mean-based approach. The most convenient way to achieve this is by constructing centered model matrix, $\bX^\star$, by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.

@@

\begin{align}\label{eq:eta_mean}
\eta^\star(x_f, \nset{{\bar{x}^\star}}) &= \hat{\beta}_f x_f + \sum \nset{\hat{\beta}} \nset{{\bar{x}^\star}} \\
\hat{y}_f  &= g^{-1} \left(\eta^\star(x_f, \nset{{\bar{x}^\star}})\right)
\end{align}
where $\nset{{\bar{x}^\star}}$ and $x_f$ are columns of $\bX^\star$ corresponding to the non-focal predictors and the focal predictor, respectively.

@@

Higher order terms such as interactions, splines, polynomials, etc., can be within the focal predictor, within (between) non-focal predictor(s) or between the focal and non-focal predictor(s). To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors such as age and wealthindex is

@@

In the first case, with \code{Age} as the focal predictor, with cubic polynomial interaction terms (and 3 model variables associated with it). In this case, each of the polynomial terms (model variables) of the focal predictor are evaluated independently across the chosen levels of the focal predictor, $Age_i$. Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section. In this case, the predictions associated with \code{Age} on the linear predictor scale become

@@

In the second case, with $Wealthindex$ as the focal predictor, the higher order terms in non-focal predictor, \code{Age}, are treated simply as additional columns in the variable space (model variables) and an appropriate choice of reference point applies just like in the models without higher order interactions. For instance, in mean-based approach, we average all non-focal model variables. Thus

@@

Lastly, consider model shown in Equation~\ref{eq:simple_inter_higher}, previously used to demonstrate how to handle interactions in non-focal predictors. Here, we use the same model to demonstrate how to handle interactions between the focal predictor, $x_2$, and non-focal predictor, $x_3$. In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points. However, for the interacting predictors, we first choose representative values for the focal predictor and some particular values of the interacting predictor similar to the way we choose values for the focal predictor or predetermined. In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting predictor $x_3$, respectively. The prediction on linear predictor scale is given by

\begin{align*}
\eta^\star(x_{2i}, x_{3j}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
\end{align*}

In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal predictor, interaction between non-focal predictors and interaction between focal and non-focal predictors.

@@

We describe the uncertainty around the estimates using confidence intervals (CI). In principle, every value of focal predictor has a different CI. The conventional way to compute variances for predictions is $\boldsymbol\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution. This generates prediction CIs and incorporates all the uncertainties -- including the uncertainty due to the intercept and non-focal predictors.  But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?

Commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal predictors when computing the CIs. Currently a non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$. We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and more complicated when the predictors are categorical. We describe our proposed method which is based on the model center and does not require input predictors to be centered prior to model fitting.

\subsection{Model center based}

Let $\bxo$ be a model matrix containing appropriately chosen values of focal predictor and centered non-focal model variables, and let $\ba$ be the anchor matrix, with same dimension and entries in all non-focal model variables as $\bxo$. Let $\baf$ be the column of $\ba$ corresponding to focal predictor model variable(s) defined in $\bxo$. Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor model variable(s). Thus the variances for the predictions we defined in the previous section is

%
\begin{align}\label{eq:centered_variance}
\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top)
\end{align}
%
We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_f^2 = \boldsymbol{0}$. Similarly, for all values of $\bxof$ close to anchor point $\bafc$ the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$. This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower (isolated) around the anchor or crosses at the anchor point for simple models.

The computation of $\bxo - \bafc$ impacts only on the intercepts and non-focal model variables, i.e., the slopes and variance corresponding to the focal predictors are not affected. This means that we can still generate effects without necessarily centering the predictors prior to model fitting.

>>>>

\item \textbf{Focal predictors:} All columns of the model matrix associated with the focal input variable. For example, all the terms of a cubic polynomial focal variable or columns corresponding to a categorical focal variable with more than two levels.
\item \textbf{Non-focal predictors:} All columns of the model matrix associated with the non-focal input variable(s). For example, in the previous household size example, all the columns associated with religion are non-focal predictors.

@@

\item \textbf{Anchor:} The value chosen for the focal input variable when estimating effect confidence intervals. The anchor choice does not affect the main prediction estimates. Typically chosen as the center point of the predictor variables corresponding to the focal input variable but other sensible values can be chosen too. If we are trying to focus on effects, the anchor from which we calculate CIs becomes important. This is not true for the more conventional prediction CIs. We can choose any anchor, e.g., zero-anchored (the value of the anchor is $0$), but center point seems like a natural, stable choice, we call this center-anchored. In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal input variable.

@@

To illustrate mean-based approach, suppose we are interested in predictions at the levels of a particular focal input variable, $x_f$, from a multivariate model. As a starting point, assume that the model has no interaction terms. The idea is to choose a reference point for the values of non-focal input variables. For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors. We refer to this as centered or mean-based approach. The most convenient way to achieve this is by constructing centered model matrix, $\bX^\star$, by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.

@@

\begin{align}\label{eq:eta_mean}
\eta^\star(x^\star_f, \nset{{\bar{x}^\star}}) &= \hat{\beta}_f x^\star_f + \sum \nset{\hat{\beta}} \nset{{\bar{x}^\star}} \\
\hat{y}_f  &= g^{-1} \left(\eta^\star(x^\star_f, \nset{{\bar{x}^\star}})\right)
\end{align}
where $\nset{{\bar{x}^\star}}$ and $x^\star_f$ are columns of $\bX^\star$ corresponding to the non-focal predictors and the focal predictor, respectively.

@@

Higher order terms such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s). To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors such as age and wealthindex is

@@

In the first case, with \code{Age} as the focal input variable, with cubic polynomial interaction ($3$ focal predictors associated with it). In this case, each of the focal predictors ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are evaluated independently across the chosen levels of the focal input variable, $Age_i$. Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section. In this case, the predictions associated with \code{Age} on the linear predictor scale become

@@

In the second case, with \code{Wealthindex} as the focal input variable, the non-focal variable, \code{Age}, is a cubic polynomial. In this case, the non-focal predictors ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are simply treated as additional columns in the predictor space and an appropriate choice of reference point applies just like in the models without higher non-focal variable interactions. For instance, in mean-based approach, we average all non-focal predictors. Thus

@@

Lastly, consider model described by Equation~\ref{eq:simple_inter_higher}, previously used to demonstrate how to handle interactions in non-focal input variables. Here, we use the same model to demonstrate how to handle interactions between the focal input variable, $x_2$, and non-focal input variable, $x_3$. In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points. However, for the interacting predictors, we first choose representative values for the focal input variable and some particular values of the interacting non-focal input variable in a similar way we choose values for the focal input variable or predetermined. In our example, suppose we pick $i$ and $j$ unique values of the focal input variable, $x_2$ and interacting input variable $x_3$, respectively. The prediction on linear predictor scale is given by

\begin{align*}
\eta^\star(x_{2i}, x_{3j}, {\bar{x}^\star_1}) = \beta_0 + \beta_1 \bar{x}^\star_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
\end{align*}

In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal input variables, interaction between non-focal input variables and interaction between focal and non-focal input variables.

@@

We describe the uncertainty around the estimates using confidence intervals (CIs). In principle, every value of focal input variable has a different CI. The conventional way to compute variances for predictions is 
%
\begin{align}\label{eq:conventional_variance}
\boldsymbol\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top}), 
\end{align}
so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution \citep{lenth2018package, fox2009effect}. This generates conventional CIs which incorporate all the uncertainties -- including the uncertainties due to the intercept and non-focal input variables.  But what if we are interested in the uncertainty as a result of the focal input variable only, so that the CIs are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?

Commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs. Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$. We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical. An alternative is our proposed approach which uses the model center and does not require the input variables to be centered prior to model fitting.

Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries in all non-focal predictors as $\bxo$. Let $\baf$ be the column of $\ba$ corresponding to focal predictor(s) defined in $\bxo$. Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor(s). Thus 
%
\begin{align}\label{eq:centered_variance}
\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top).
\end{align}
%
We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_f^2 = \boldsymbol{0}$. Similarly, for all values of $\bxof$ close to anchor point $\bafc$, the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$. This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models. This is the effect of the particular focal input variable. By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs, Equation~\ref{eq:conventional_variance}.

The computation of $\bxo - \bafc$ impacts only on the intercepts and non-focal predictors, i.e., the slopes and variance corresponding to the focal predictors are not affected. This means that we can still generate effects without necessarily centering the predictors prior to model fitting.
